---
title: "Reproducible Research Assignment I"
author: "Harish"
date: "Friday, August 14, 2015"
output: html_document
---

#Assignment I
This is the first assignment in the Reproducible Research course.
This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

##Data
The data for this assignment is a dataset containing three variables
 steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
 date: The date on which the measurement was taken in YYYY-MM-DD format
 interval: Identifier for the 5-minute interval in which measurement was taken
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset

##Goal of the Assignment
To complete the entire assignment in a single R markdown document that can be processed by knitr and be transformed into an HTML file.


###Loading and preprocessing the data
```{r echo=TRUE}

if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")) {
        temp <- tempfile()
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
        unzip(temp)
        unlink(temp)
}

data <- read.csv("activity.csv")
```

###What is mean total number of steps taken per day?


```{r echo=TRUE}
steps_by_day <- aggregate(steps ~ date, data, sum)
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")


rmean <- mean(steps_by_day$steps)
rmedian <- median(steps_by_day$steps)


steps_by_interval <- aggregate(steps ~ interval, data, mean)

plot(steps_by_interval$interval,steps_by_interval$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")

```

The total number of steps taken per day is 
```{r} 
steps_by_day 
```

The mean is 
```{r echo=TRUE} 
rmean 
``` 
and the median is 
```{r echo=TRUE} 
rmedian 
```


###What is the average daily activity pattern?

 Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

 Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r echo=TRUE}
steps_by_interval <- aggregate(steps ~ interval, data, mean)

plot(steps_by_interval$interval,steps_by_interval$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")


max_interval <- steps_by_interval[which.max(steps_by_interval$steps),1]

```

The 5-minute interval, on average across all the days in the data set, containing the maximum number of steps is 
```{r echo=TRUE} 
max_interval 
```


###Imputing missing values

Missing data needed to be imputed. Only a simple imputation approach was required for this assignment. Missing values were imputed by inserting the average for each interval. Thus, if interval 10 was missing on 10-02-2012, the average for that interval for all days (0.1320755), replaced the NA.

```{r echo=TRUE}
incomplete <- sum(!complete.cases(data))
imputed_data <- transform(data, steps = ifelse(is.na(data$steps), steps_by_interval$steps[match(data$interval, steps_by_interval, steps_by_interval$interval)], data$steps))
```

Zeroes were imputed for 10-01-2012 because it was the first day and would have been over 9,000 steps higher than the following day, which had only 126 steps. NAs then were assumed to be zeros to fit the rising trend of the data. 

```{r echo=TRUE}

imputed_data[as.character(imputed_data$date) == "2012-10-01", 1] <- 0

```


###Recount total steps by day and create Histogram. 

```{r echo=TRUE}

steps_by_day_i <- aggregate(steps ~ date, imputed_data, sum)
hist(steps_by_day_i$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")

#Create Histogram to show difference. 
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="red", xlab="Number of Steps", add=T)
legend("topright", c("Imputed", "Non-imputed"), col=c("blue", "red"), lwd=10)

```

###Calculate new mean and median for imputed data. 
```{r echo=TRUE}
rmean.i <- mean(steps_by_day_i$steps)
rmedian.i <- median(steps_by_day_i$steps)
```

###Calculate difference between imputed and non-imputed data.
```{r echo=TRUE}
mean_diff <- rmean.i - rmean
med_diff <- rmedian.i - rmedian
```

###Calculate total difference.
```{r echo=TRUE}
total_diff <- sum(steps_by_day_i$steps) - sum(steps_by_day$steps)
```

1.The imputed data mean is 
```{r echo=TRUE}
rmean.i
```
2.The imputed data median is 
```{r echo=TRUE}
rmedian.i
```
3.The difference between the non-imputed mean and imputed mean is  
```{r echo=TRUE}
mean_diff
```
4.The difference between the non-imputed median and imputed median is  
```{r echo=TRUE}
med_diff
```



###Are there differences in activity patterns between weekdays and weekends?

Created a plot to compare and contrast number of steps between the week and weekend. There is a higher peak earlier on weekdays, and more overall activity on weekends. 
```{r echo=TRUE}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", 
              "Friday")
imputed_data$dow = as.factor(ifelse(is.element(weekdays(as.Date(imputed_data$date)),weekdays), "Weekday", "Weekend"))

steps_by_interval_i <- aggregate(steps ~ interval + dow, imputed_data, mean)

library(lattice)

xyplot(steps_by_interval_i$steps ~ steps_by_interval_i$interval|steps_by_interval_i$dow, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")
```



###Observation:
Are there differences in activity patterns between weekdays and weekends? Yes. The plot indicates that the person moves around more (or more active) during the weekend days.




